---
title: "Initial data exploration"
output: html_notebook
---

This notebook contains code and output from exploring the data to inform cleaning steps and analysis process.

```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(GGally)
library(psych)
```

# Load and inspect data

## Daily energy data

Dataset 1, from https://www.kaggle.com/datasets/emmanuelfwerr/london-homes-energy-data.

```{r}
daily_energy <- read_csv("../3_raw_data/london_energy.csv") %>% 
  janitor::clean_names()
```

```{r}
skimr::skim(daily_energy)
```

No missing values in kaggle aggregated energy data

### Averaging daily energy consumption (across all households)

Not sure if mean is appropriate measure here, let's look at histogram of energy consumption values of every household on a single date:

```{r}
daily_energy %>% 
  select(date)
```

```{r}
# boxplot on 2012-12-31
daily_energy %>%
  filter(date == "2012-12-31") %>% 
  ggplot(aes(x = kwh)) + 
  geom_boxplot()

# histogram on 2012-12-31
daily_energy %>%
  filter(date == "2012-12-31") %>% 
  ggplot() + 
  geom_histogram(aes(x = kwh), bins = 150)
```

Energy consumption on an individual date, frequency of households, is not normally distributed but right-skewed. Median may be a better average daily measure than mean.


### Energy consumption patterns over all households

```{r}
# median
daily_energy %>% 
  group_by(date) %>% 
  mutate(median_kwh = median(kwh)) %>% 
  ggplot(aes(x = date, y = median_kwh)) +
  geom_line()
```


```{r}
# mean
daily_energy %>% 
  group_by(date) %>% 
  mutate(mean_kwh = mean(kwh)) %>% 
  ggplot(aes(x = date, y = mean_kwh)) +
  geom_line()
```

Definitely some seasonality in energy consumption, whichever average used.


### Energy consumption patterns within households

Let's look at a few individual households:

```{r}
daily_energy %>% 
  distinct(lc_lid)
```

```{r}
daily_energy %>% 
  filter(lc_lid %in% c("MAC000002", "MAC000058", "MAC000359", "MAC000714", "MAC000933", "MAC000997")) %>% 
  ggplot(aes(x = date, y = kwh, colour = lc_lid)) +
  geom_line() +
  facet_wrap(~lc_lid)
```

Can already see households are behaving differently, some high consumers, others less so; some change over the year, some less so.

Might want to use a rolling average here, e.g. 7-day, to give a smoother line.

```{r}
# tbc rolling average
```

### types of household (by consumption patterns)

look to see if there are obvious high/medium/low consumer groups

```{r}
household_alltime_consumption <- daily_energy %>% 
  group_by(lc_lid) %>% 
  summarise(alltime_median_kwh = median(kwh),
            alltime_mean_kwh = mean(kwh))
```

```{r}
household_alltime_consumption
```

```{r}
household_alltime_consumption %>% 
  ggplot() +
  geom_histogram(aes(x = alltime_median_kwh), bins = 200)
```

Look at january or july dates only, when there may be more variation in usage (cold and hot weather)

```{r}
household_jan_jul_consumption <- daily_energy %>% 
  mutate(month = month(date, label = TRUE, abbr = FALSE)) %>% 
  filter(month %in% c("January","July")) %>% 
  group_by(lc_lid, month) %>% 
  summarise(median_kwh = median(kwh),
            mean_kwh = mean(kwh))
```

```{r}
# january medians
household_jan_jul_consumption %>% 
  filter(month == "January") %>% 
  ggplot() +
  geom_histogram(aes(x = median_kwh), bins = 200)

# july medians
household_jan_jul_consumption %>% 
  filter(month == "July") %>% 
  ggplot() +
  geom_histogram(aes(x = median_kwh), bins = 200)
```

Most households have median daily consumption in range 0-20 kWh in both January and July.

```{r}
household_jan_jul_consumption %>% 
  ggplot() +
  geom_boxplot(aes(x = month, y = median_kwh))
```

Maybe energy consumption in January is slightly more than in July. Could test this. Long tail of high energy consumers.

No obvious groups to split into, but could label households above Q3 as "high consumers", below Q1 as "low consumers", and rest as "medium consumers".

```{r}
all_kwh_stats <- daily_energy %>% 
  select(kwh) %>% 
  skimr::skim()

all_kwh_stats %>% 
  colnames()

kwh_q1 <- all_kwh_stats %>% 
  select(numeric.p25) %>% 
  pull()

kwh_q3 <- all_kwh_stats %>% 
  select(numeric.p75) %>% 
  pull()

kwh_q1
kwh_q3
```
Note these are calculated from all dates, not adjusted for jan/july/seasonality and note the date range covers more winters (3) than summers (2).

```{r}
# add household type, can be used as lookup key-value (join with full data)
household_alltime_consumption <- household_alltime_consumption %>% 
  mutate(household_consumption_level = case_when(
    alltime_median_kwh >= kwh_q3 ~ "high",
    alltime_median_kwh > kwh_q1 ~ "average",
    alltime_median_kwh <= kwh_q1 ~ "low",
    .default = NA_character_
  ),
  household_consumption_level = factor(household_consumption_level, levels = c("low", "average", "high")))
```

```{r}
household_alltime_consumption %>% 
  ggplot() +
  geom_boxplot(aes(x = household_consumption_level, y = alltime_median_kwh))
```

```{r}
household_alltime_consumption %>% 
  summarise(count = n(), .by = household_consumption_level)
```

Seems like a potentially useful split of data

```{r}
daily_energy_processed <- left_join(daily_energy, household_alltime_consumption, by = "lc_lid")

daily_energy_processed
```

Look at median energy over time by type:

```{r}
daily_energy_processed_medians <- daily_energy_processed %>% 
  group_by(date, household_consumption_level) %>% 
  summarise(median_kwh = median(kwh)) 

daily_energy_processed_medians %>% 
  arrange(date)

daily_energy_processed_medians %>% 
  ggplot() +
  geom_line(aes(x = date, y = median_kwh, colour = household_consumption_level))
```

There may be a way to group households by consumption in an unsupervised way, i.e. by clustering instead.. tbc

## Daily weather data

Dataset 2, from https://www.kaggle.com/datasets/emmanuelfwerr/london-weather-data.



Ideas:

* types of household - look to see if there are any groups by patterns of consumption (not just overall high/low but variable v fixed over time) - try clustering?
* types of weather days - see if there are clusters of temperature, precipitation, cloud cover, sunshine etc --> categories "mild cloudy rainy day" etc
* convert datetime to season, month, day, weekday


