---
title: "Ts forecasting model for household energy consumption"
output: html_notebook
---

# Setup

```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(tsibble) # to convert to tsibble (ts data)
library(slider) # for rolling average
library(feasts) # for ts analyses
library(fable) # for forecasting
library(urca) # support function for ARIMA model?
#library(GGally) # for ggpairs to explore correlations
#library(psych) # for pairplots to explore correlations
```

```{r}
# household energy consumption by date (from Kaggle)
daily_energy <- read_csv("../3_raw_data/london_energy.csv") %>% 
  janitor::clean_names()
```
```{r}
# trim dates - because of lack of data at start & unusual end value
  # start in Winter 2011 (91 households by that point)
  # and remove last date, very low median value (unusual)
daily_energy_trim <- daily_energy %>% 
  filter(date >= "2011-12-01" & date < "2014-02-28")

daily_energy_trim
```
```{r}
total_days <- as.numeric(max(daily_energy_trim$date) - min(daily_energy_trim$date) + 1)
total_days
```

```{r}
daily_energy_trim %>% 
  summarise(num_obs = n(), .by = lc_lid)
```

5,566 households in this data, varying number of observations each over total time frame of 820 days

```{r}
# generate household stats
hhold_stats <- daily_energy_trim %>% 
  mutate(zero_kwh = if_else(kwh == 0, T, F)) %>% 
  group_by(lc_lid) %>%
  summarise(num_values = n(),
            num_zeros = sum(zero_kwh),
            min_value = min(kwh),
            max_value = max(kwh),
            range = (max(kwh) - min(kwh)),
            median_kwh = median(kwh),
            iqr = IQR(kwh),
            mean_kwh = mean(kwh),
            sd = sd(kwh)
            ) %>% 
  mutate(pc_missing_days = (100 * (total_days - num_values) / total_days), .after = lc_lid) %>% 
  mutate(pc_zero_values = (100 * num_zeros / num_values), .after = pc_missing_days) %>% 
  ungroup()

hhold_stats
```

```{r}
daily_energy_clean %>%
  filter(lc_lid == "MAC000002") %>% 
  ggplot() +
  geom_line(aes(x = date, y = kwh)) +
  ggtitle("Daily energy consumption for a single household (MAC000002)") +
  ylab("Energy consumed (kWh)") + xlab("Date")
```

``` {r}
# list hholds with more than 10% of data as 0 kWh
hholds_10pc_zero <- hhold_stats %>% 
  filter(pc_zero_values > 10) %>% 
  pull(lc_lid) # 73 hholds

# list hholds with fewer than 180 observations (6 months) in dataset
hholds_180obs <- hhold_stats %>% 
  filter(!lc_lid %in% hholds_10pc_zero) %>% 
  filter(num_values < 180) %>% 
  pull(lc_lid) # 19 hholds

# create cleaned df (remove hholds with limited data or lots of zeros, trim dates)
daily_energy_clean <- daily_energy_trim %>% 
  filter(!lc_lid %in% hholds_10pc_zero) %>% 
  filter(!lc_lid %in% hholds_180obs)

num_hholds_og <- length(unique(daily_energy$lc_lid))
num_hholds_clean <- length(unique(daily_energy_clean$lc_lid))

num_hholds_og
num_hholds_clean
num_hholds_og - num_hholds_clean
100* (num_hholds_og - num_hholds_clean) / num_hholds_og
```

```{r}
skimr::skim(daily_energy_clean)
```

After cleaning: 92 households (1.6%) removed from data, leaving 5,474 households in "cleaned" data, across 820 days from 2011-12-01 to 2014-02-27, with varying numbers of households per day (note on-ramp at the beginning of the time period).

```{r}
# generate tsibble from cleaned data
daily_energy_ts <- as_tsibble(daily_energy_clean, key = lc_lid, index = date)
```


# Prelim

Brief: Using all households (n = 5,474; after cleaning), model expected average household consumption in future years.

## How to do test/train split?

Depends on amount of data!

In the first 6 months or so, not every household signed up:

```{r}
num_hholds_by_date <- daily_energy_clean %>% 
  group_by(date) %>% 
  summarise(num_hholds = n())

num_hholds_by_date %>% 
  ggplot() +
  geom_line(aes(x = date, y = num_hholds))
```

So there might not be enough data in early timeframe, or long enough time period, to do a split.

Initial: try training on ~1.5 years data, test on the most recent winter months, e.g.

* Training: 2011-12-01 to 2013-10-31
* Test: 2013-11-01 to 2014-02-27

```{r}
start_date <- min(daily_energy_clean$date)
end_date <- max(daily_energy_clean$date)

train_length <- as.numeric(as_date("2013-10-31") - start_date + 1) # +1 inclusive dates
test_length <- as.numeric(end_date - as_date("2013-11-01") + 1)
pc_test <- 100 * test_length / (train_length + test_length)

train_length
test_length
# should sum to 820!
pc_test
```

So this would mean forecasting for winter 2013, ~15% of the timeframe of the actual dataset, covering one winter season.

What would be the % data?

```{r}
test_data_len <- daily_energy_clean %>% 
  filter(date >= "2013-11-01") %>%  # the test split
  nrow()
test_data_len

pc_test_data <- 100* test_data_len / nrow(daily_energy_clean)
pc_test_data
```

601,038 observations in test set, 17.3% of overall observations across whole dataset.

So 2013-11-01 (as first date in test data) seems like an appropriate cut-off date for test/train split.

```{r}
# split the data

# ~85% of the timeframe, just under 2 years of data
allhholds_forecast_train <- daily_energy_ts %>% 
  filter_index(. ~ "2013-10-31")

# most recent 4 months
allhholds_forecast_test <- daily_energy_ts %>% 
  filter_index("2013-11-01" ~ .)
```


### Other options to explore if this doesn't work

_Q: can I test/train split the most recent year to have some training data but test on 20% households?_ If so, how to pick the 20%? Could do random selection, but would I be missing a key stratification (e.g. high/medium/low consumers - could calculate from each household's median: high >Q3 overall, low <Q1 overall, medium in Q1-Q3 overall range)

## How/what to model on?

From exploration:

* For a traditional forecast (1 value per day) - use median as average per day and per household due to not normal distribution?
* Or use all data available and do probabilistic forecasting -- see work by Jethro Browell

* Visually, see seasonal pattern (high winter, low summer), weekday pattern (higher weekends that weekdays), and looks like energy consumption is reducing over time. -- use seasonal models:
  * snaive
  * SARIMA
  * seasonal decomposition + any model - tbc


```{r}
allhholds_forecast_train
```

Use ```autoplot()``` to plot ts graph - but reduce rows first, otherwise R gets stuck!

__PAUSING__ to pursue a different business question (characterising households by patterns of energy consumption)

